Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-10-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

TODO: send complete manuscript by september 30th!

@@

The research objective of the popmis dataset is to predict the pupils' popularity based on their gender and the experience of the teacher.  The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction. The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`:

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
```{r ignore}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred)
```

@@

```{r impact-pred, include=FALSE, eval=FALSE}
pred <- quickpred(dat)
plot_pred(pred)

meth <- make.method(dat)
meth[meth != ""] <- "2l.pmm"
```

[TODO: mutate data to get the right data types for imputation (e.g. integer for clustering variable).]

```{r imp_impact, eval = FALSE, include=FALSE}
dat <- dat %>% mutate(across(everything(), as.integer))
imp <- mice(dat, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
```

```{r impact-fit, eval = FALSE, include=FALSE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

```{r eval=FALSE, include=FALSE}
meth <- make.method(hiv)
meth["hiv"] <- "2l.heckman"
imp <- mice(hiv, method = meth)
pred <- make.predictorMatrix(hiv)
pred[, "region"] <- -2
pred[, "smoke"] <- -3
imp2 <- mice(hiv, method = meth, predictorMatrix = pred)
```

```{r}
data("hiv",  package = "GJRM")
# We select  5 predictor variables over 9 regions
colnames(hiv)
hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
# Study/group variable has to be recoded as integer
hivdata$region<-as.integer(hivdata$region)

@@

hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))

summary(hivdata)
#  We obtain here the random effects for each interviewer, this is an approximation 
#of the interviewer's skill which will be used as an exclusion constraint. 
#Here, since the location of the interviewer was not randomly assigned to the 
#subjects, the assignment was corrected for region and language.

hivdata$hivconsent <- as.factor(hivdata$hivconsent)
ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
reffect  <- ranef(ID_mixed)$interviewerID
reffect$interviewerID <- levels(hivdata$interviewerID)
colnames(reffect)     <- c("IDreffect","interviewerID")
hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)

hivdata$interviewerID<-NULL
hivdata$hivconsent<-NULL
hivdata$language<-NULL
summary(hivdata)

# #Set prediction matrix and methods
# ini <- mice(hivdata, maxit = 0)
# meth<-ini$method
# meth["hiv"]<-"2l.binary"
# pred <- ini$pred
# pred[,"region"] <- 0
# pred["region",] <- 0
# pred["hiv","region"]<- -2
# pred["hiv","IDreffect"]<- 0
# 
# # MAR imputation model, we used the 2l.binary from the miceadds package
# data_imp_mar <- mice( data= hivdata, # dataset with missing values
#                       m = 10,   # number of imputations
#                       meth = meth, #imputation method vector
#                       pred = pred, #imputation predictors matrix
#                       maxit=1)
# 
# # source("mice.impute.2l.heckman.R")
# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["1"]]))
# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["1"]]))
# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["2"]]))
# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["2"]]))
# 
# # Heckman model
# pred["hiv","IDreffect"] <- -3
# meth<-ini$method
# meth["hiv"]<-"2l.heckman"
# 
# data_imp_heckp <- mice( hivdata, # dataset with missing values
#                         m = 10,   # number of imputations
#                         seed = 1234, #seed attached to the dataID
#                         meth = meth, #imputation method vector
#                         pred = pred, #imputation predictors matrix
#                         print = T,
#                         meta_method="reml",
#                         pmm=FALSE)

#Set prediction matrix and methods
ini <- mice(hivdata, maxit = 0)
meth<-ini$method
meth["hiv"]<-"2l.binary"
pred <- ini$pred
pred[,"region"] <- 0
pred["region",] <- 0
pred["hiv","region"]<- -2
pred["hiv","IDreffect"]<- 0

# MAR imputation model, we used the 2l.binary from the miceadds package
data_imp_mar <- mice( data= hivdata, # dataset with missing values
                      m = 20,   # number of imputations
                      meth = meth, #imputation method vector
                      pred = pred, #imputation predictors matrix
                      maxit=1)

source("mice.impute.2l.heckman.R")

# Heckman model
pred["hiv","IDreffect"] <- -3
meth<-ini$method
meth["hiv"]<-"2l.heckman"

data_imp_heckp <- mice( hivdata, # dataset with missing values
                        m = 20,   # number of imputations
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        maxit=1,
                        meta_method="reml",
                        pmm=FALSE)
```

@@

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. See also all of the TODOs that need to be worked out. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

@@

The research objective of the `popmis` dataset is to predict the pupils' popularity based on their gender and the experience of the teacher.  The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction. The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`:

@@

<!-- --- -->
Mutate data to get the right data types for imputation (e.g. integer for clustering variable).

```{r}
dat <- dat %>% mutate(across(everything(), as.integer))
```

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
```{r impact}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred)

meth <- make.method(dat)
meth
```

@@

Impute the incomplete data
```{r imp_impact}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

```{r impact-fit}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

@@

Load the data:
```{r}
data("hiv",  package = "GJRM")
# We select  5 predictor variables over 9 regions
colnames(hiv)
hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
```

Recode variables:
```{r}
# Study/group variable has to be recoded as integer
hivdata$region<-as.integer(hivdata$region)

@@

hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
```

We obtain here the random effects for each interviewer, this is an approximation of the interviewer's skill which will be used as an exclusion constraint. Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.
```{r}
hivdata$hivconsent <- as.factor(hivdata$hivconsent)
ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
reffect  <- ranef(ID_mixed)$interviewerID
reffect$interviewerID <- levels(hivdata$interviewerID)
colnames(reffect)     <- c("IDreffect","interviewerID")
hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)

hivdata$interviewerID<-NULL
hivdata$hivconsent<-NULL
hivdata$language<-NULL
```

Set the Heckman model as imputation method
```{r}
#Set prediction matrix and methods
ini <- mice(hivdata, maxit = 0)
meth<-ini$method
meth["hiv"]<-"2l.binary"
pred <- ini$pred
pred[,"region"] <- 0
pred["region",] <- 0
pred["hiv","region"]<- -2
pred["hiv","IDreffect"]<- 0

# Heckman model
pred["hiv","IDreffect"] <- -3
meth<-ini$method
meth["hiv"]<-"2l.heckman"
```

Impute the data:
```{r}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

@@

- Adding evaluations of the imputations such as convergence checks

### 2022-10-14 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
version show work in progress/under construction. The last section is
not part of the manuscript, but purely for reminders. We aim to submit
at JSS, so there is no word count limit (``There is no page limit, nor a
limit on the number of figures or tables''). {[}Just adding some text to
get a better guess of what the actura abstract will look like: Lorem
ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor
incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam,
quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo
consequat. Duis aute irure dolor in reprehenderit in voluptate velit
esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat
cupidatat non proident, sunt in culpa qui officia deserunt mollit anim
id est laborum.{]}
}

@@

The research objective of the popmis dataset is to predict the pupils'
popularity based on their gender and the experience of the teacher. The
analysis model corresponding to this dataset is multilevel regression
with random intercepts, random slopes and a cross-level interaction. The
outcome variable is \texttt{popular}, which is predicted from the
unit-level variable \texttt{sex} and the cluster-level variable
\texttt{texp}:

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.820     0.353    13.648     4.245     0.000    33.200     0.979 
sex             0.971     0.376     2.581     4.128     0.059    63.141     0.989

                            Estimate 
Intercept~~Intercept|school    0.259 
Residual~~Residual             1.065 
ICC|school                     0.197

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.121     0.390    13.125     4.242     0.000    33.503     0.979 
sex             0.673     0.418     1.611     4.121     0.180    66.343     0.989

                            Estimate 
Intercept~~Intercept|school    0.317 
Residual~~Residual             1.253 
ICC|school                     0.213

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.941     0.422    11.708     4.203     0.000    39.844     0.982 
sex             1.029     0.537     1.916     4.076     0.127   105.671     0.993

                            Estimate 
Intercept~~Intercept|school    0.306 
Residual~~Residual             1.308 
ICC|school                     0.195

@@

Create a methods vector and predictor matrix, and make sure
\texttt{name} is not included as predictor:

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/ignore-1} \end{center}

@@

{[}TODO: mutate data to get the right data types for imputation
(e.g.~integer for clustering variable).{]}

>>>>

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
version show work in progress/under construction. The last section is
not part of the manuscript, but purely for reminders. See also all of
the TODOs that need to be worked out. We aim to submit at JSS, so there
is no word count limit (``There is no page limit, nor a limit on the
number of figures or tables''). {[}Just adding some text to get a better
guess of what the actura abstract will look like: Lorem ipsum dolor sit
amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut
labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud
exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
Duis aute irure dolor in reprehenderit in voluptate velit esse cillum
dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
proident, sunt in culpa qui officia deserunt mollit anim id est
laborum.{]}
}

@@

The research objective of the \texttt{popmis} dataset is to predict the
pupils' popularity based on their gender and the experience of the
teacher. The analysis model corresponding to this dataset is multilevel
regression with random intercepts, random slopes and a cross-level
interaction. The outcome variable is \texttt{popular}, which is
predicted from the unit-level variable \texttt{sex} and the
cluster-level variable \texttt{texp}:

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.932     0.274    18.012     4.420     0.000    19.514     0.964 
sex             0.859     0.284     3.026     4.216     0.036    37.601     0.981

                            Estimate 
Intercept~~Intercept|school    0.265 
Residual~~Residual             1.007 
ICC|school                     0.213

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.054     0.266    19.029     4.554     0.000    14.936     0.954 
sex             0.732     0.306     2.393     4.233     0.071    34.886     0.980

                            Estimate 
Intercept~~Intercept|school    0.318 
Residual~~Residual             1.257 
ICC|school                     0.208

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.714     0.193    24.432     5.032     0.000     8.224     0.919 
sex             1.286     0.288     4.460     4.218     0.010    37.215     0.981

                            Estimate 
Intercept~~Intercept|school    0.299 
Residual~~Residual             1.050 
ICC|school                     0.228

@@

Mutate data to get the right data types for imputation (e.g.~integer for
clustering variable).

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- dat %>% mutate(across(everything(), as.integer))
\end{CodeInput}
\end{CodeChunk}

Create a methods vector and predictor matrix, and make sure
\texttt{name} is not included as predictor, but as clustering variable:

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}

\begin{CodeInput}
R> pred[pred == 1] <- 2
R> pred["mort", ] <- 2
R> pred[, "mort"] <- 2
R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
R> pred[, "name"] <- -2
R> diag(pred) <- 0
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-2} \end{center}

\begin{CodeInput}
R> meth <- make.method(dat)
R> meth
\end{CodeInput}
\begin{CodeOutput}
       name        type         age motor_score       pupil          ct 
         ""          ""          ""          ""       "pmm"       "pmm" 
     hypots        tsah         edh        mort 
      "pmm"       "pmm"       "pmm"          "" 
\end{CodeOutput}
\end{CodeChunk}

Impute the incomplete data

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.33813790 0.340459131  -6.867602 6.978862e-12
2                    type -0.41509929 0.179936773  -2.306918 2.107792e-02
3                     age  0.03034813 0.001561834  19.431086 0.000000e+00
4 as.factor(motor_score)2 -0.66052409 0.068846376  -9.594174 0.000000e+00
5 as.factor(motor_score)3 -1.05030912 0.070059114 -14.991756 0.000000e+00
6 as.factor(motor_score)4 -1.49637069 0.071492961 -20.930322 0.000000e+00
7                   pupil  0.48922641 0.038536483  12.695149 0.000000e+00
8                      ct  0.42634415 0.028388807  15.018037 0.000000e+00
             b          df dfcom          fmi       lambda m          riv
1 9.395965e-04  8668.10341 11013 0.0099557149 0.0097273074 5 0.0098228573
2 1.729929e-05 10991.51028 11013 0.0008229571 0.0006411648 5 0.0006415761
3 8.332782e-09 10482.95010 11013 0.0042891736 0.0040992241 5 0.0041160969
4 5.649577e-05  6979.24803 11013 0.0145856016 0.0143032580 5 0.0145108098
5 3.003051e-05  9526.87220 11013 0.0075503299 0.0073420043 5 0.0073963080
6 3.329522e-05  9362.42789 11013 0.0080288174 0.0078169354 5 0.0078785213
7 3.337070e-04    54.63795 11013 0.2949938851 0.2696512493 5 0.3692088868
8 1.180517e-05  5893.23780 11013 0.0179108151 0.0175775781 5 0.0178920775
          ubar
1 1.147849e-01
2 3.235648e-02
3 2.429325e-06
4 4.672029e-03
5 4.872243e-03
6 5.071289e-03
7 1.084612e-03
8 7.917581e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10522.944 10588.713 -5252.472 10504.944     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2913  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32713                 -0.41953                  0.03031  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65651                 -1.04719                 -1.49327  
                  pupil                       ct  
                0.48990                  0.42335  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10525.723 10591.492 -5253.861 10507.723     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2937  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.30150                 -0.41940                  0.03024  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65497                 -1.04451                 -1.49454  
                  pupil                       ct  
                0.47281                  0.42888  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10512.28 10578.05 -5247.14 10494.28    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2852  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35320                 -0.41028                  0.03034  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66933                 -1.05474                 -1.50379  
                  pupil                       ct  
                0.49491                  0.42442  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10530.02 10595.79 -5256.01 10512.02    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2923  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32670                 -0.41286                  0.03037  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66801                 -1.04770                 -1.50070  
                  pupil                       ct  
                0.47210                  0.43109  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10488.615 10554.383 -5235.307 10470.615     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2939  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.38215                 -0.41343                  0.03049  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65379                 -1.05740                 -1.48956  
                  pupil                       ct  
                0.51641                  0.42397  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

attr(,"class")
[1] "mitml.result" "list"        
\end{CodeOutput}
\begin{CodeInput}
R> # testEstimates(as.mitml.result(fit))
\end{CodeInput}
\end{CodeChunk}

\hypertarget{case-study-iii-hiv-data}{%
\section{Case study III: HIV data}\label{case-study-iii-hiv-data}}

@@

R=region+language+(1\textbar InterviewID) model with with interviewer as
random effects, because the observations are not independent. Interviews
are not allocated randomly. In theory we expect the
inclusion-restriction variable to be randomly assigned, that's why we're
adding region and language to compensate for non-random allocation.

Load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("hiv",  package = "GJRM")
R> # We select  5 predictor variables over 9 regions
R> colnames(hiv)
\end{CodeInput}
\begin{CodeOutput}
 [1] "hivconsent"      "hiv"             "age"             "education"      
 [5] "wealth"          "region"          "marital"         "std"            
 [9] "age1sex_cat"     "highhiv"         "partner"         "condom"         
[13] "aidscare"        "knowsdiedofaids" "evertestedHIV"   "smoke"          
[17] "religion"        "ethnicity"       "language"        "interviewerID"  
[21] "sw"             
\end{CodeOutput}
\begin{CodeInput}
R> hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
\end{CodeInput}
\end{CodeChunk}

Recode variables:

\begin{CodeChunk}
\begin{CodeInput}
R> # Study/group variable has to be recoded as integer
R> hivdata$region<-as.integer(hivdata$region)
R> 
R> # Categorical variables have to be recoded as factor
R> # to use the 2l.binary imputation method, it is required that the level names should not have any space between words
R> 
R> hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
R> #the missing variable is stored as a factor in the dataset, otherwise the method will 
R> #apply the imputation correction for a  missing continuous variable instead for 
R> #a missing binary variable,which is in this case the binary response of the hiv test.
R> 
R> hivdata$marital <- as.factor(hivdata$marital)
R> levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
R> hivdata$condom <- as.factor(hivdata$condom)
R> levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
R> hivdata$highhiv <- as.factor(hivdata$highhiv)
R> levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
R> hivdata$interviewerID <- as.factor(hivdata$interviewerID)
R> 
R> hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
R> interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
\end{CodeInput}
\end{CodeChunk}

We obtain here the random effects for each interviewer, this is an
approximation of the interviewer's skill which will be used as an
exclusion constraint. Here, since the location of the interviewer was
not randomly assigned to the subjects, the assignment was corrected for
region and language.

\begin{CodeChunk}
\begin{CodeInput}
R> hivdata$hivconsent <- as.factor(hivdata$hivconsent)
R> ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
R> reffect  <- ranef(ID_mixed)$interviewerID
R> reffect$interviewerID <- levels(hivdata$interviewerID)
R> colnames(reffect)     <- c("IDreffect","interviewerID")
R> hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
R> 
R> hivdata$interviewerID<-NULL
R> hivdata$hivconsent<-NULL
R> hivdata$language<-NULL
\end{CodeInput}
\end{CodeChunk}

Set the Heckman model as imputation method

\begin{CodeChunk}
\begin{CodeInput}
R> #Set prediction matrix and methods
R> ini <- mice(hivdata, maxit = 0)
R> meth<-ini$method
R> meth["hiv"]<-"2l.binary"
R> pred <- ini$pred
R> pred[,"region"] <- 0
R> pred["region",] <- 0
R> pred["hiv","region"]<- -2
R> pred["hiv","IDreffect"]<- 0
R> 
R> # Heckman model
R> pred["hiv","IDreffect"] <- -3
R> meth<-ini$method
R> meth["hiv"]<-"2l.heckman"
\end{CodeInput}
\end{CodeChunk}

Impute the data:

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice( hivdata, # dataset with missing values
+                         m = 10,   # number of imputations
+                         maxit = 1,
+                         seed = 1234, #seed attached to the dataID
+                         meth = meth, #imputation method vector
+                         pred = pred, #imputation predictors matrix
+                         print = T,
+                         meta_method="reml",
+                         pmm=FALSE)
\end{CodeInput}
\begin{CodeOutput}

 iter imp variable
  1   1  hiv
  1   2  hiv
  1   3  hiv
  1   4  hiv
  1   5  hiv
  1   6  hiv
  1   7  hiv
  1   8  hiv
  1   9  hiv
  1   10  hiv
\end{CodeOutput}
\end{CodeChunk}
